Index measures for oak decline severity using phenotypic descriptors

### 2020-02-06 manuscript/manuscript.Rmd
proofing and finished alterations from Nathans comments

The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model response surfaces for specific example cases.
The same example cases were used as described for the LIME analyses above.
This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation [@abbasi2012].

For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from the supervised random forest regression model, building a model response surface.

@@

The same approach was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
The upper boundary for the ranges of active bleed size and black stain size was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.

@@

### Site differences

A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b). 
To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in the Materials and Methods.
After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.

@@

The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
Other tree size related descriptors such as and live crown ratio stem bleeding descriptors contributed secondarily.

The DAI was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure 2b).
The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to the DAI index model.
These were followed closely by the stem staining and bleeding descriptors such as bleed prevalence. 
The presence of *A. biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')`^th^.

@@

LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table 1).
These descriptors were also the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
Increasing decline severity was associated with a reduction in the size and condition of the crown.
For instance, declining trees were found to have a composite crown volume below 841m^3^.
Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severely declining trees a condition below 23.9%.

For the DAI, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 2).
This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.
The important descriptors identified were mainly associated with tree size such as live crown ratio and lower crown radius.
Bleed prevalence and stem symptom descriptors were not found to be locally important to any of the decline type cases.
Live crown ratio greater than 61.6% in the severe AOD and moderate COD cases.
COD trees were found to have greater crown contact (greater than 50%) compared to AOD trees.

@@

The complex descriptor interactions within the index models can be investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models. 
The index values generated by the models from these simulated phenotypic data can then be visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted. 
Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.

Crown condition descriptors such as composite crown volume and crown condition contributed highly to the PDI (Figure 2a).
The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
It can be seen in Figure 3 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and composite crown volume decrease, the PDI increases. 
A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g. greater crown radius, total height, lower crown height and DBH; Figures 3a & b), with relatively small stature trees having PDIs of 0.6 or greater even with 100% crown condition.

The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
Simulated response surfaces for the DAI are shown in Figure 4. 
These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively). 
Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
Also higher DAI values could be obtained from trees with greater stature (Figure 4c). 
Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD however only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.

@@

Crown condition and volume were shown to be the key contributors to the PDI. 
This is to be expected and an indicator of model suitability for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007]. 
However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
It has been shown here that in our visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).

The high contribution of bleed prevalence related descriptors to the DAI is also expected as this is one of the defining features of AOD [@denman2014]. 
There was no contribution of *A. biguttatus* exit hole occurrence with with the DAI and therefore stem bleeding. This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previously that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A. biguttatus* larvae complete their life cycle [@brown2017a;@vansteenkiste2004].

AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).
The association of crown condition and AOD severity could not previously be shown by @denman2014 due to a low sample size.

@@

The remission trees as a group were shown to be intermediate in decline severity and type to the healthy, AOD and COD trees on both the PDI and DAI index scales.
However, this was also the most varied of the decline status groups.
It is possible that this could reflect an underpin relationship of the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
Using these decline indexes, remission trees can be more objectively defined in terms of their temporal trajectory within the decline index continuum.

@@

An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual decline status classification in the field [@ferretti1997].
This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g.. COD trees with DAI values > 0.5).
The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification and still allow the data driving flexibility of the unsupervised approach.

>>>>

The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model ($M_{PDI}$,$M_{DAI}$) response surfaces for specific example cases.
The same example cases were used as described for the LIME analyses above.
This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation [@abbasi2012].

For $M_{PDI}$, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from $M_{PDI}$, building a model response surface.

@@

The same approach was used for $M_{DAI}$ using a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
The upper boundary for the ranges of active bleed size and black stain size was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.

@@

### Phenotypic differences between site locations

A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b). 
To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in the Materials and Methods.
After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the adjustment was able to remove the majority of the between site variance.

@@

The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
Other tree size related descriptors such as and live crown ratio stem bleeding descriptors contributed secondarily.

$M_{DAI}$ was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure 2b).
The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to $M_{DAI}$.
These were followed closely by the stem staining and bleeding descriptors such as bleed prevalence. 
The presence of *A. biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')`th.

@@

LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases using lime (Table 1).
These descriptors were also the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
Increasing decline severity was associated with a reduction in the size and condition of the crown.
For instance, declining trees were found to have a composite crown volume below 841m^3^.
Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severely declining trees a condition below 23.9%.

For $M_{DAI}$, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 2).
This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.
The important descriptors identified were mainly associated with tree size such as live crown ratio and lower crown radius.
Bleed prevalence and stem symptom descriptors were not found to be locally important to any of the decline type cases.
Live crown ratio greater than 61.6% in the severe AOD and moderate COD cases.
COD trees were found to have greater crown contact (greater than 50%) compared to AOD trees.

@@

The complex descriptor interactions within the index models ($M_{PDI}$,$M_{DAI}$) investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models. 
The index values generated by the models from these simulated phenotypic data were then visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted. 
Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.

Crown condition descriptors such as composite crown volume and crown condition contributed highly to $M_{PDI}$ (Figure 2a).
The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
It can be seen in Figure 3 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and composite crown volume decrease, the PDI increases. 
A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g. greater crown radius, total height, lower crown height and DBH; Figures 3a & b), with relatively small stature trees having PDIs of 0.6 or greater even with 100% crown condition.

The DAI ($M_{DAI}$) is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
Simulated response surfaces for the DAI are shown in Figure 4. 
These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively). 
Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
Higher DAI values could also be obtained from trees with greater stature (Figure 4c). 
Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD however only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.

@@

Crown condition and volume were shown to be the key contributors to the $M_{PDI}$. 
This is to be expected and are an indicator of model suitability for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007]. 
However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
It has been shown here that in our visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).

The high contribution of bleed prevalence related descriptors to $M_{DAI}$ is also expected as this is one of the defining features of AOD [@denman2014]. 
There was no contribution of *A. biguttatus* exit hole occurrence with with the DAI and therefore stem bleeding. This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previously that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A. biguttatus* larvae complete their life cycle [@brown2017a;@vansteenkiste2004].

AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).
The association of crown condition and AOD severity has previously be shown by @brown2016.

@@

The remission trees as a group were shown to be intermediate in decline severity and type to the healthy, AOD and COD trees on both the PDI and DAI index scales.
However, this was also the most varied of the decline status groups.
It is possible that this could reflect the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
Using these decline indexes, remission trees could be more objectively defined in terms of their temporal trajectory within the decline index continuum.

@@

An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual decline status classification in the field [@ferretti1997].
This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g. COD trees with DAI values > 0.5).
The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification and still allow the data driven flexibility of the unsupervised approach.
